Coordinated monitoring of the Piney Point wastewater discharge into Tampa Bay: Data synthesis and reporting

### 2022-05-18 manu-draft.Rmd
finished dashboard setup text

The local workflow created by TBEP was hosted online in a GitHub repository [https://github.com/tbep-tech/piney-point, @Beck21d] that allowed use of additional resources that were critical to the synthesis effort (Figure \@ref(fig:synthesis), testing and sharing).  Hosting the RStudio project on GitHub served multiple purposes of 1) using version control to document and track changes using Git, 2) making the code publicly accessible for transparency with the larger research and management community, 3) allowing automated testing of repository content, and 4) serving as an access point for deploying the dashboard on Shiny Server (see below).  The use of automated testing was an important service to identify potential errors or mistakes in the synthesized datasets. The testthat R package [@Wickham11] was used to write several checks for each dataset that were considered critical for maintaining integrity of the informaiton.  Each time changes were made to the repository and pushed to GitHub, tests were run automatically with GitHub Actions to verify each dataset included appropriate and accurate results.  For example, a test was used to verify that all water quality parameters were appropriately named and did not include any missing values. A notification was sent via email to TBEP staff if an error was encountered.  This automated service reduced the amount of time required to manually check the data and provided assurance that the reported results in the dashboard were correct.

@@

The Piney Point environmental monitoring dashboard was created to communicate the results from the synthesized datasets.  The dashboard was developed "on the fly" as data became available and the synthesis workflow described in the previous section was used to update input datasets during the course of the 2021 monitoring effort.  All code used to synthesize the datasets from Google Drive and the code used to create the dashboard were contained in the same repository, allowing the workflow in Figure \@ref(fig:synthesis) to be used iteratively to update the content.  The dashboard was created using the shiny [@Chang21] and flexdashboard [@Iannone20] R packages to create the server and user interface components of the app. The flexdashboard package was particularly useful in creating an intuitive user interface to organize related components of the monitoring effort. The dashboard was deployed using Shiny Server by pulling the GitHub repository to a remote server maintained by TBEP each time an update was made (Figure \@ref(fig:synthesis), app deployment).  This allowed anyone to access the dashboard using the url https://shiny.tbep.org/piney-point.

Every shiny application includes a server and user interface.  The server components of the dashboard include

* App creation and hosting, Figure \@ref(fig:dashboard), @Beck21c
* App structure - backend
     * R/dat_proc.R to data folder, logs
     * Testing why and how (testthat, GitHub Actions)
     * Rmarkdown w/ flexdashboard: Dependencies/globals, reactives, UI setup with text descriptions and dynamic inputs
     * CSS styling
     * Google Analytics js
* App structure - frontend
     * Landing page
     * Current data
          * Water quality results
          * Algal surveys
          * Seagrass/macroalgae
          * Contaminants
          * Benthic
     * Baseline data
          * Water quality
          * Seagrasses

@@

* Analytics and where it was linked (e.g., DEP, myfwc), note analytics did not start until May 13th, 2021
     * Average time spent on app, mobile vs desktop (go to analytics to get this)
     * Total users over time (from Google Analytics), Figure \@ref(fig:analytics)
     * Users by geographic location (country, state, city), Table \@ref(tab:demtab)

>>>>

The local workflow created by TBEP was hosted online in a GitHub repository [https://github.com/tbep-tech/piney-point, @Beck21d] that allowed use of additional resources that were critical to the synthesis effort (Figure \@ref(fig:synthesis), testing and sharing).  Hosting the RStudio project on GitHub served multiple purposes of 1) using version control to document and track changes using Git, 2) making the code publicly accessible for transparency with the larger research and management community, 3) allowing automated testing of repository content, and 4) serving as an access point for deploying the dashboard on Shiny Server (see below).  The use of automated testing was an important service to identify potential errors or mistakes in the synthesized datasets. The testthat R package [@Wickham11] was used to write several checks for each dataset that were considered essential for maintaining integrity of the information.  Each time changes were made to the repository and pushed to GitHub, tests were run automatically with GitHub Actions to verify each dataset included appropriate and accurate results.  For example, a test was used to verify that all water quality parameters were appropriately named and did not include any missing values. A notification was sent via email to TBEP staff if an error was encountered.  This automated service reduced the amount of time required to manually check the data and provided assurance that the reported results in the dashboard were correct.

@@

The Piney Point environmental monitoring dashboard was created to communicate the results from the synthesized datasets [Figure \@ref(fig:dashboard), @Beck21c].  The dashboard was developed "on the fly" as data became available and the synthesis workflow described in the previous section was used to update input datasets during the course of the 2021 monitoring effort.  All code used to synthesize the datasets from Google Drive and the code used to create the dashboard were contained in the same repository, allowing the workflow in Figure \@ref(fig:synthesis) to be used iteratively to update the content.  The dashboard was created using the shiny [@Chang21] and flexdashboard [@Iannone20] R packages to create the server and user interface components of the app. The flexdashboard package was particularly useful in creating an intuitive user interface to organize related components of the monitoring effort. The dashboard was deployed using Shiny Server by pulling the GitHub repository to a remote server maintained by TBEP each time an update was made (Figure \@ref(fig:synthesis), app deployment).  This allowed anyone to access the dashboard using the url https://shiny.tbep.org/piney-point.

```{r}
# dashboard code breakdown
nlns <- 2346
setlns <- 346
rctlns <- 1319
uilns <- nlns - (setlns + rctlns)
srvlns <- setlns + rctlns
setpr <- paste0(round(100 * setlns / nlns, 0), '%')
rctpr <- paste0(round(100 * rctlns / nlns, 0), '%')
uipr <- paste0(round(100 * uilns / nlns, 0), '%')
srvpr <- paste0(round(100 * srvlns / nlns, 0), '%')
```

Both the server and user interface (UI) components of the dashboard were included in a single R Markdown [@Xie20] file totaling `r nlns` lines of code, with `r srvpr` and `r uipr` of the code dedicated to the server and UI, respectively.  The server components of the dashboard included a setup component (`r setlns` lines, `r setpr` of the total) and reactive component (`r rctlns`, `r rctpr` of the total).  The setup imported all R package dependencies (15 total), the synthesized input data from Google Drive (20 \.RData files), an R script with custom functions to summarize or plot the results, five log files, and various "static" objects used in other parts of the dashboard that did not depend on user inputs. The rest of the server included thirty-one reactive objects (`r rctlns` lines, `r rctpr` of the total) that received different user inputs and returned a shiny data object sent back to the user interface.  The imported log files were created during the data synthesis steps external to the dashboard and were specific to each data type (e.g., water quality, phytoplankton, etc.).  Each log file was used to display when a data type was last changed, to inform users when updates were made to the input data.

The remainder of the R Markdown file was used to create the UI or the dashboard (`r uilns` lines, `r uipr` of total).  The flexdashboard package rendered with R Markdown was used to place shiny content in specific CSS elements, such as boxes or tabs that are scaled automatically to a user's browser, to create a UI with easily navigable content.  Specifically, the UI code organized the dashboard content into 1) a landing page when a user first accesses the website (Figure \@ref(fig:dashboard)), 2) tabs for navigating to current results for data from a specific monitoring type, and 3) tabs for viewing historical baseline data for comparison with the response-based monitoring data.  The landing page included a text overview of the dashboard, a map showing the monitoring locations, and several summary boxes at the bottom showing current effort to date (e.g., number of sites sampled, total days sampling, etc., Figure \). Each tab for the current data was organized specifically for each monitoring data type, although common elements included maps and tabular summaries where the displayed data could be chosen by date ranges or selected parameters.  Finally, the baseline data tab displayed long-term monitoring data for water quality and seagrass to provide a comparison with the response-based monitoring data.

An important service provided by the dashboard was creating context for the response-based monitoring results relative to baseline conditions in Tampa Bay.  This allowed dashboard users to quickly assess if the 2021 conditions were abnormal relative to typical seasonal patterns observed in the bay.  In addition to the baseline data tabs, all water quality data from the response-based monitoring were compared to the long-term monthly averages for data at the nearest ambient monitoring station.  Elements of the UI and summary plots allowed a user to quickly determine if the results were outside of the expected range.  Specifically, map points for a sampling event were displayed with a solid outline if results were out of range and without an outline if within the normal range.  A user could also click on an individual map point to view a time series of the 2021 data for the selected location that included upper and lower limits for the long-term monthly averages.  This functionality required additional modifications to the UI, in addition to developing custom code in the synthesis workflow that summarized the normal ranges prior to uploading data to the dashboard.  Ultimately, these changes improved the quality of information provided by the response-based monitoring by leveraging the decades of long-term monitoring data available for the bay.

@@

* Google Analytics js
* Analytics and where it was linked (e.g., DEP, myfwc), note analytics did not start until May 13th, 2021
     * Average time spent on app, mobile vs desktop (go to analytics to get this)
     * Total users over time (from Google Analytics), Figure \@ref(fig:analytics)
     * Users by geographic location (country, state, city), Table \@ref(tab:demtab)
